---
title: "Elongation"
author: "Sara V. Ramalhete/Isabel Duarte"
date: "2024-10-31"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install and load required packages

Install the required packages (RUN ONLY ONCE - Comment after first usage)

```{r}
install.packages("ggplot2")
install.packages("ggthemes")
install.packages("FactoMineR")
install.packages("factoextra")
install.packages('GGally')
install.packages("magrittr")
install.packages("dplyr")
install.packages("tidyverse")
install.packages("viridis")
install.packages("ggridges")
```

Load Required packages

```{r}

library(ggplot2)
library(ggthemes)
library(FactoMineR)
library(factoextra)
library(GGally)
library(magrittr)
library(dplyr)
library(tidyverse)
library(viridis)
library(ggridges)
```

Import the DataSet

```{r}

db_original_REF <- read.csv("./Ref_RealTime.csv", sep=";", dec=",")
```

## 1C

Scatterplot + lineplot for each measure (hue by embryo)

```{r}

db_original_REF$Embryo <- as.factor(db_original_REF$Embryo)


medidas <- function(a, b){
  
  # Creating the plot
  zz <- ggplot(data = db_original_REF, aes(x = Time, y = a, group = Embryo, color = Embryo)) +
        geom_hline(yintercept = seq(0, 8, 1), color = "gray94") +
    geom_vline(xintercept = seq(-10, 25, 5), color = "gray94") +
    geom_line(linewidth = 0.7) + 
    geom_point(size = 1) +
    theme_minimal() +
    ggtitle(b) +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      axis.title = element_text(size = 11),
      axis.text = element_text(size = 10, colour = "black"),
      axis.line = element_line(color = "black"),
      axis.ticks = element_line(),
      legend.text = element_text(size = 11),
      legend.title = element_blank(),
      legend.key.size = unit(13, "point"),
      ) +scale_x_continuous(breaks = c(-10, -5, 0, 5, 10, 15, 20, 25))+
    xlab("Time (h)") + 
    ylab("Elongation (mm)")
  
  return(zz)
}



p1 <- medidas(db_original_REF$C.pPL, 'C-pPL') + guides(color='none') + theme(axis.title.x = element_blank())

p2 <- medidas(db_original_REF$C.PS, 'C-PS') + guides(color='none')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p3 <- medidas(db_original_REF$C.N, 'C-N') + guides(color='none')+ theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p4 <- medidas(db_original_REF$PSM, 'PSM') + guides(color='none') + theme(axis.title.x = element_blank())

p5 <- medidas(db_original_REF$SEG, 'SEG') + guides(color='none')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p6 <- medidas(db_original_REF$C.HF, 'C-HF')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p7 <- medidas(db_original_REF$C.Seg, 'C-Seg') + guides(color='none')

p8 <- medidas(db_original_REF$N.pPL, 'N-pPL') + guides(color='none')+theme(axis.title.y = element_blank())

p9 <- medidas(db_original_REF$PS, 'PS') + guides(color='none')+theme(axis.title.y = element_blank())


svg(filename="Crescimento.svg",
    width=12, 
    height=10, 
    pointsize=14)


patchwork::wrap_plots(p1, p2, p3,
                      p4, p5, p6,
                      p7, p8, p9, ncol = 3)

dev.off()
```

## 1F

Normalized measures by max

```{r}

db_Normalized <- db_original_REF

names <- colnames(db_Normalized)[3:11]

for (i in names) {
  
  for (j in 1:nrow(db_Normalized)) {
    
    db_Normalized[j,i] <- db_Normalized[j,i]/max(db_Normalized[,i], na.rm = T)
    
  }
  
}

medidas_nor <- function(x, y){
  
  zz <- ggplot(data=db_Normalized, aes(x=Time, y=x, group=Embryo))+ 
    geom_point(size=4, color="#C8C8C8", alpha=0.6)+ 
    geom_smooth(aes(group = 1), method = "lm", linetype = "solid", color = "black", se = FALSE)+
    theme_minimal() +
    ggtitle(y)+
    theme(plot.title = element_text(hjust = 0.5, size = 14),
          axis.title = element_text(size = 11),
          axis.text = element_text(size = 10, colour = "black"),
          axis.line = element_line(color = "black"),
          legend.text = element_text(size = 11),
          legend.title = element_blank()) +
    xlab("Time (h)")+ ylab("Elongation (a.u.)")+
    expand_limits(x=c(-10,26), y=c(0,1)) +
    scale_x_continuous(breaks = c(-10, -5, 0, 5, 10, 15, 20, 25))+
    scale_y_continuous(breaks = c(0.00, 0.25, 0.50, 0.75, 1.00))
  
  return(zz)
  
}


p1_n <- medidas_nor(db_Normalized$C.pPL, 'C-pPL')+theme(axis.title.x = element_blank())

p2_n <- medidas_nor(db_Normalized$C.PS, 'C-PS')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p3_n <- medidas_nor(db_Normalized$C.N, 'C-N')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p4_n <- medidas_nor(db_Normalized$PSM, 'PSM')+theme(axis.title.x = element_blank())

p5_n <- medidas_nor(db_Normalized$SEG, 'SEG')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p6_n <- medidas_nor(db_Normalized$C.HF, 'C-HF')+theme(axis.title.x = element_blank(), axis.title.y = element_blank())

p7_n <- medidas_nor(db_Normalized$C.Seg, 'C-Seg')

p8_n <- medidas_nor(db_Normalized$N.pPL, 'N-pPL')+theme(axis.title.y = element_blank())

p9_n <- medidas_nor(db_Normalized$PS, 'PS')+theme(axis.title.y = element_blank())


svg(filename="Crecimento_Norm.svg", 
    width=12, 
    height=10, 
    pointsize=14)

patchwork::wrap_plots(p1_n, p2_n, p3_n,
                      p4_n, p5_n, p6_n,
                      p7_n, p8_n, p9_n, ncol = 3)

dev.off()
```

## 3A

PCA analyze

```{r}

output_pca <- PCA(db_original_REF[,colnames(db_original_REF) %in% names], scale.unit = TRUE, ncp = 5, graph = TRUE)

svg(filename="PCA.svg", 
    width=10, 
    height=10, 
    pointsize=14)

fviz_pca_var(output_pca, col.var = "black", repel=T)

dev.off()


```

## 3B

Correlation plot for each measure and time

```{r}

db_original_REF$Embryo <- as.character(db_original_REF$Embryo)

column_labels <- c("Time(h)", "C-pPL(mm)", "C-PS(mm)", "C-N(mm)", "N-pPL(mm)", "N-PS(mm)", "PSM(mm)", "C-Seg(mm)", "SEG(mm)", "C-HF(mm)")

gcor <- ggpairs(db_original_REF,
        columns = 2:11,
        columnLabels = column_labels,
        aes(color=Embryo),
        lower = list(continuous = wrap("smooth", size=1, se = F)),
        upper  = list(continuous = "blank"),
        diag  = list(continuous = "blankDiag"),
        axisLabels = c("show", "internal", "none"),
        switch = "both") +
  theme(strip.placement = "outside",
        strip.background =element_blank(),
        axis.text.x = element_text(size = 7),
        axis.title.x = element_text(size = 8),
        axis.text.y = element_text(size = 7),
        axis.title.y = element_text(size = 8))

svg(filename="Correlation.svg", 
    width=20, 
    height=20, 
    pointsize=14)

gcor

dev.off()

```

## 1D

Prepare the function to return the elongation rate

```{r}

taxas_vector <- function(x){
  
  db_rate <- db_original_REF[,which(colnames(db_original_REF) %in% c("Time", "Embryo", x))]

db_rate <- na.omit(db_rate)

db_rate <- split(db_rate, db_rate$Embryo)

db_rate <- db_rate[sapply(db_rate, nrow) > 2]

for (i in 1:length(db_rate)) {
  db_rate[[i]] <- db_rate[[i]][which(db_rate[[i]]$Time==min(db_rate[[i]]$Time, na.rm = T) | db_rate[[i]]$Time==max(db_rate[[i]]$Time, na.rm = T)),]
  }



for (i in 1:length(db_rate)) {
  
  for (j in 1:nrow(db_rate[[i]])) {
    
    db_rate[[i]]$Teste[j] <- ifelse(db_rate[[i]]$Time[j]==min(db_rate[[i]]$Time, na.rm = T), "Min", "Max")
    
}
}



Taxa <- vector()

for (i in 1:length(db_rate)) {
  
  Taxa[i] <- (db_rate[[i]][which(db_rate[[i]]$Teste=="Max"),x] - db_rate[[i]][which(db_rate[[i]]$Teste=="Min"),x]) / (db_rate[[i]][which(db_rate[[i]]$Teste=="Max"),"Time"] - db_rate[[i]][which(db_rate[[i]]$Teste=="Min"),"Time"])
  
}

return(Taxa)

}
```

Setup the dataframe of elongation rate to each measure

```{r}

Final_taxa <- list()
length(Final_taxa) <- 9
names(Final_taxa) <- colnames(db_original_REF)[3:11]

Tabela <- NULL
valores <- NULL

for (i in names(Final_taxa)) {
  
  Final_taxa[[i]] <- as.data.frame(taxas_vector(i))
  Final_taxa[[i]][2] <- i
  colnames(Final_taxa[[i]]) <- c("Taxa", "Measure")
  
  c <- Final_taxa[[i]]
  
  Tabela <- rbind(c, Tabela)
  
  Quadro <- data.frame(Measure= i,
                       media = mean(Final_taxa[[i]]$Taxa),
                       desvio = sd(Final_taxa[[i]]$Taxa))
  
  valores <- rbind(Quadro, valores)
}

```

Plot elongation rate

```{r}

desired_order <- c('C.pPL', 'C.PS', 'C.N', 'PSM', 'SEG', 'C.HF', 'C.Seg', 'N.pPL', 'N.PS')

valores$Measure <- factor(valores$Measure, levels = desired_order)

library(ggplot2)

gtax <- ggplot(valores, aes(x=Measure, y=media, fill=Measure))+
  geom_bar(stat="identity")+
  scale_fill_manual(values = c("C.pPL" = "#3F6E9A", "C.PS" = "#9A68A4", "C.N" = "#F4A652", "PSM" = "#FFD0D2", 'SEG' = '#91FFFF', 'C.HF' = '#B59FEB', 'C.Seg' = '#FFE0B3', 'N.pPL' = '#7ABC6D', 'N.PS' = '#2A883E'))+
  geom_errorbar(aes(ymin=media-desvio, ymax=media+desvio), width=0.4, colour="black", alpha=0.9)+
  geom_text(aes(label = round(media, 2)), vjust = ifelse(valores$media<0, 1.5, -0.5), color = "black") +
  theme_minimal()+
  scale_x_discrete(labels = c("C.pPL" = "C-pPL", "C.PS" = "C-PS", "C.N" = "C-N", "PSM" = "PSM", 'SEG' = 'SEG', 'C.HF' = 'C-HF', 'C.Seg' = 'C-Seg', 'N.pPL' = 'N-pPL', 'N.PS' = 'N-PS')) +
  theme(legend.position="none",
        axis.text.x= element_text(colour = "black", size=11))+
  scale_y_continuous(name="Elongation Rate (mm/h)", limits=c(-0.15, 0.3))+
  labs(x = NULL)


svg(filename="Elongation_rate.svg", 
    width=8, 
    height=6, 
    pointsize=14)

gtax

dev.off()
```

## 2C

Format dataset

```{r}

my.hh.order <- c("HH4","HH5","HH6","HH7","HH8","HH9","HH10")

db_original_REF %>%
  select(Embryo, HHStage, Time, C.pPL, C.PS, C.N, N.pPL, N.PS, PSM, C.Seg, SEG, C.HF) %>%
  filter(!is.na(HHStage)) %>%
  mutate_at(vars(HHStage), funs(paste0("HH", .))) %>%
  mutate(HHStage = factor(HHStage, levels = my.hh.order)) -> clean.embryo.length

clean.embryo.length %>% 
    group_by(Embryo) -> embryo.length.by.embryo

clean.embryo.length %>%
  group_by(HHStage) -> embryo.length.by.HHStage

# summary stats
count(embryo.length.by.embryo)
count(embryo.length.by.HHStage)
```

Create the ridge line plots for LENGTH (all 9 measures)

```{r}

for (i in 1:9) {
  my.xlabs <- c("1.Total embryo length (mm)", "2.Total PS length (mm)", 
                "3.Notochord length (mm)", "4.Posterior length (mm)", 
                "5.PS length (mm)", "6.PSM length (mm)", "7.Anterior length (mm)", 
                "8.Segmented length (mm)", "9.Head fold length (mm)")
  plot <-
    ggplot(clean.embryo.length, mapping=aes(x = clean.embryo.length[,i+3], y = HHStage, fill = HHStage)) +
    geom_density_ridges(alpha=0.75, rel_min_height = 0.001, scale = 2,
                        quantile_lines = TRUE, quantiles=2,
                        jittered_points = TRUE, position = "points_sina",
                        point_alpha = 0.8) +
    theme_ridges(grid = TRUE, center_axis_labels = TRUE)+
    theme(legend.position = "none") +
    xlab(my.xlabs[i]) +
    ylab('HH stage') +
    ggtitle("Embryo length distribution per HH stage")
  print(plot)
  ggsave(filename=paste0("ridgeplot_overlap",i,".pdf"),
         width = 8.27, height = 5.83, units = "in",
         device = "pdf", path="output/")
}
```

## 3B

Create the ridge line plots for TIME

```{r}

plot.time <-
    ggplot(clean.embryo.length, mapping=aes(x = Time, y = HHStage, fill = HHStage)) +
    geom_density_ridges(alpha=0.75, rel_min_height = 0.001, scale = 1,
                        quantile_lines = TRUE, quantiles=2,
                        jittered_points = TRUE, position = "points_sina",
                        point_alpha = 0.8) +
    theme_ridges(grid = TRUE, center_axis_labels = TRUE)+
    theme(legend.position = "none") +
    scale_x_continuous(breaks = seq(-10, 25, 5)) +
    xlab("Time (h)") +
    ylab('HH stage') +
    ggtitle("Embryo time distribution per HH stage")
print(plot.time)
ggsave(plot.time,filename=paste0("ridgeplot_time",i,".pdf"),
       width = 8.27, height = 5.83, units = "in",
       device = "pdf", path="output/")
```
